\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{hyperref}

\usepackage{minted}
\author{Alexandre Moine}
\title{Roadmap to lint\_ligo 2.0}
\begin{document}
\maketitle

This document lists several directions for the future of
\verb|lint_ligo|\footnote{\url{https://github.com/nobrakal/lint_ligo}},
a linter for LIGO\footnote{\url{https://ligolang.org/}}.

\paragraph{Integration in integrated development environments}

A linter is even more useful when integrated into an IDE, as linting
errors could be displayed directly to the user.
%
An integrated \verb|lint_ligo| could be a serious argument for the
adoption of the linter by the community.
This can be achieved in a standard way by implementing a Language
Server Protocol\footnote{\url{https://microsoft.github.io/language-server-protocol/}}.

\paragraph{A standard set of linting rules}

The linter provides an efficient pattern-matching algorithm but it
needs a concrete set of patterns to be useful. Some patterns are
available in the \verb|examples/| directory but there are too few of
them. Code patterns should be added to create a concrete standard set
of linting rules.

\paragraph{Semantic patterns}
Some code patterns target more the code semantics than its exact
shape. For example, the three definitions of the following
\verb|option| function in CameLIGO:

\begin{minted}{ocaml}
let option (default : int) (x : int option) : int =
  match x with
  | None -> default
  | Some x -> x

let option (default : int) (x : int option) : int =
  match x with
  | Some x -> x
  | None -> default

let option (default : int) (x : int option) : int =
  match x with
    None -> default (* Note the absence of a leading vertical bar. *)
  | Some x -> x
\end{minted}

have the exact same semantics and should be matched by the same
semantic pattern if needed. This is not what happens in the current
implementation since the actual pattern-matching algorithm is based on
the CST, the first structure of the compilation pipeline. This
structure still contains many textual information irrelevant for a
semantic pattern (like the order of the pattern-matching cases or the
leading vertical bar).

An improvement could be to normalize the CST by a compilation followed
by a decompilation before running the pattern-matching
algorithm. Irrelevant information is erased during the first stages of
the compilation and the decompilation allows us to run the existing
algorithm.

\paragraph{A better detection of unused variables}

The current algorithm used for detecting unused variables returns an
under-approximation of the full set of unused variables. Indeed, it
only detects if each variable is being used, but does not propagate
the information.

For example with the program:
\begin{minted}{ocaml}
  let x = 42 in
  let y = x  in
  ()
\end{minted}

Only the variable \verb|y| will be marked as unused because
\verb|x| is used in the definition of \verb|y|.

A possible improvement is to use a refined algorithm propagating the
information: in the previous example, the algorithm should be aware
that, as \verb|y| is unused and \verb|x| is only used in its
definition, \verb|x| is finally not used.

\paragraph{Detection of unused types}

For now, the unused variables algorithm ignores type definitions. It
could be interesting to verify if locally defined types are actually
used in the code.

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
